\change{The effect of the COVID-19 pandemic on arXiv pre-prints in maths, physics, computer science, and quantitative biology}

### 2022-10-27 article/ms.tex
revviewer comments

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics are posting more papers on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in general physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\noindent Peer-reviewed publications are the most common measure of productivity in \change{many specialties of science and technology}. The \arxiv\cite{Ginsparg:2011} ({https://arxiv.org}) is a service for the distribution of research publications (pre-prints) prior to their appearance in a journal. The presence of a pre-print on \arxiv\ does not ensure that the contents have passed peer-review, but most material on \arxiv\ eventually goes through peer-review because it is now standard in many research specialties to post to \arxiv\ either during or after the peer-review process \cite{Lariviere:2014}. \change{The fraction of publications that are also posted to \arxiv\ increases each year but varies significantly by specialty, from 18.5\% in mathematics to nearly 100\% in high energy physics\cite{?,??}. 
Not all academic outputs (peer-reviewed publications or otherwise) in a given research area will appear on \arxiv, but in the specialties we investigate here (physics, mathematics, and other STEM-related fields) the number of pre-prints posted on \arxiv\ is proportional to (a proxy measure of) the number of peer-reviwed publications in that area. For this reason, even if the proportionality differs between specialities of research, we can still examine changes per specialty as a function of time before and during the COVID-19 pandemic}.

@@

We retrieved metadata for 1,660,423 pre-prints posted to \arxiv\ between 1 April 2007 and 31 May 2022. The metadata includes the creation date, research field(s), title, author name(s), abstract, and other miscellaneous information\cite{Clement:2019}. There is an increasing number of pre-prints posted to \arxiv\ each year in nearly every field (Figure~\ref{fig:1}). These long-term trends are relatively predictable from year to year, allowing us to quantify any change in academic productivity due to the COVID-19 pandemic. The number of pre-prints posted varies substantially by category. Computer science (\texttt{cs}) is the growing the fastest with more than 5,000 pre-prints per month, followed by math (\texttt{math}) with about 3,500 per month. For contrast, only $\approx$200 pre-prints are posted each month quantitative biology (\texttt{q-bio}), and just $\approx$40 in nuclear experimental physics (\texttt{nucl-ex}). For this reason, we model the number of monthly pre-prints in each research field separately. We used the number of publications from January 2015 to December 2019 to train a predictive Gaussian process model\cite{Rasmussen:2006}. This model treats the number of monthly pre-prints in a research field as a sequence of normally distributed random variables. The expected number of pre-prints was assumed to increase linearly with time, and a quasi-periodic kernel function\cite{Rasmussen:2006,Ambikasaran:2014} was used to capture seasonal variations (Figure~\ref{fig:1}). In nearly all research fields, the number of pre-prints posted since January 2020 agree closely with the model predictions (Table~1) indicating no significant overall decline in academic productivity due to the COVID-19 pandemic. %In fact, the total number of \arxiv\ pre-prints in 2020 across all fields (168,630) exceeds our model predictions ($162{,}577 \pm 4{,}393$; $+1.4\sigma$).  % todo

The exception is general physics (\arxiv\ category \texttt{physics}; not all physics-related domains), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to general physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began (Figure~\ref{fig:physics}). All other physics sub-categories remained largely stable with time. The category of mathetmatical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than physics. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study.

@@

The field of quantitative biology (\texttt{q-bio}) showed the largest increase in pre-prints in 2020 above what is expected by our model, before returning to pre-pandemic levels in mid-2021. % todo: 
There were {2,790} quantitative biology pre-prints in 2020, {50\%} above the previous year, representing a $+8\sigma$ deviation from our model predictions (predicted 1,914 $\pm$ 109; 876 fewer than observed). This increase is explainable by an increase in COVID-19 related research, as there were 864 quantitative biology pre-prints in 2020 with pandemic-related terms in their title or abstract (see Methods), and just 39 in the decade prior. Indeed, in {April 2020} nearly 60\% of the \arxiv\ pre-prints in quantitative biology were related to the pandemic (Figure~\ref{fig:2}). Pandemic-related pre-prints also appeared in other fields (computer science, physics, statistics, and quantitative finance, all peaking around {April 2020}), but these papers constituted fewer than 10\% of the total pre-prints in these fields. Approximately 10--15\% of current pre-prints in quantitative biology are related to the ongoing pandemic.This represents a strong shift in research focus, but this fraction is declining with time. % todo:

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date.
However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in {general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of publications as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

We have taken a simple approach to address name confusion. We first define a unique author by family name and the initial of the first given name (`Family-name, I.'), such that we intentionally group together authors that may share the same initial of their first given name. 
While our approach to name confusion is grossly simple, it is unlikely that these choices have any substantial impact on our inferences. In general, this errs on the side of name conflation and over-productivity, except for the relatively rare authors who change their last names over their career.
Any common name is likely to appear in the literature early in the data set and will not impact the conclusions we draw about how publishing changed from 2020 January onwards. Indeed, we re-produced the analysis and figures in this paper by taking all initials given and it made no change to our conclusions. This is due, in part, to most authors not providing middle initials.

@@

We identified a pre-print as being related to the COVID-19 pandemic if the title or abstract contained any of the following four (case-insensitive) terms: `pandemic', `COVID', `SARS-CoV-2', `lockdown', or `coronavirus'. Before 2020, these terms rarely appear in the title or abstract of any pre-print on \arxiv\ (Figure~\ref{fig:2}).

@@

\change{There were 234 pre-prints posted to the \texttt{q-bio} primary category in 2020 which include pandemic-related terms, and where the lead author had never before appeared on the \texttt{q-bio} pre-print before. We investigated every pre-print to extract the affiliation of those authors, and used this information to search the internet for the author's website. From this information we identified their research specialisations using their own words, or from their social media pages if no website could be found. In situations where an author's specialisation crossed multiple fields (e.g., mathematics and biology) we took a conservative approach and described the author as a biologist. However, we found that only 3\% (8/234) were led by established biologists who had not appeared in \texttt{q-bio} pre-prints before. About half (45\%; 107/234) were written by mathematicians or physicists, with another 26\% (61/234) led by engineers, computer scientists, or economists.}

@@

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics \change{in science and technology specialisations} are posting more \change{manuscripts} on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in general physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\noindent Peer-reviewed publications are the most common measure of productivity in \change{many specialties of science and technology}. The \arxiv\cite{Ginsparg:2011} ({https://arxiv.org}) is a service for the distribution of research manuscripts (pre-prints) prior to their appearance in a journal. The presence of a pre-print on \arxiv\ does not ensure that the contents have passed peer-review, but most material on \arxiv\ eventually goes through peer-review because it is now standard in many research specialties to post to \arxiv\ either during or after the peer-review process \cite{Lariviere:2014}. 
\change{The fraction of peer-reviewed publications that are also posted to the \arxiv\ increases each year\cite{Davis:2007} but not all academic outputs (peer reviewed publications or otherwise) will appear on the \arxiv. Fundamental reports, findings from inquiries or research groups, and many other texts that are only published in peer-reviewed publications will never appear on the \arxiv. In the specialisations we investigate here (physics, mathematics, astronomy, computer science, and other STEM-related fields), the fraction of peer-reviewed literature that also appears on \arxiv\ is estimated to vary from 18.5\% in mathematics to nearly 100\% in high energy physics\cite{Davis:2007,GentilBeccot:2009}. Although the fraction may vary between discplines, within a given specialisation the number of pre-prints provides a proxy measure of the number of peer-reviewed publications, which we can use to examine changes in each specialty as a function of time.}

@@

We retrieved metadata for 1,660,423 pre-prints posted to \arxiv\ between 1 April 2007 and 31 May 2022. The metadata includes the creation date, research field(s), title, author name(s), abstract, and other miscellaneous information\cite{Clement:2019}. There is an increasing number of pre-prints posted to \arxiv\ each year in nearly every field (Figure~\ref{fig:1}). These long-term trends are relatively predictable from year to year, allowing us to quantify any change in academic productivity due to the COVID-19 pandemic. The number of pre-prints posted varies substantially by category. Computer science (\texttt{cs}) is the growing the fastest with more than 5,000 pre-prints per month, followed by math (\texttt{math}) with about 3,500 per month. For contrast, only $\approx$200 pre-prints are posted each month quantitative biology (\texttt{q-bio}), and just $\approx$40 in nuclear experimental physics (\texttt{nucl-ex}). For this reason, we model the number of monthly pre-prints in each specialisation separately. We used the number of publications from January 2015 to December 2019 to train a predictive Gaussian process model\cite{Rasmussen:2006}. This model treats the number of monthly pre-prints in a research field as a sequence of normally distributed random variables. The expected number of pre-prints was assumed to increase linearly with time, and a quasi-periodic kernel function\cite{Rasmussen:2006,Ambikasaran:2014} was used to capture seasonal variations (Figure~\ref{fig:1}). In nearly all research fields, the number of pre-prints posted since January 2020 agree closely with the model predictions (Table~1) indicating no significant overall decline in academic productivity due to the COVID-19 pandemic. %In fact, the total number of \arxiv\ pre-prints in 2020 across all fields (168,630) exceeds our model predictions ($162{,}577 \pm 4{,}393$; $+1.4\sigma$).  % todo

The exception is general physics (\arxiv\ category \texttt{physics}; not all physics-related domains), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to general physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began (Figure~\ref{fig:physics}). All other physics sub-categories remained largely stable with time. The category of mathematical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than \change{the general} physics \change{category}. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study.

@@

The field of quantitative biology (\texttt{q-bio}) showed the largest increase in pre-prints in 2020 above what is expected by our model, before returning to pre-pandemic levels in mid-2021. % todo: 
There were {2,790} quantitative biology pre-prints in 2020, {50\%} above the previous year, representing a $+8\sigma$ deviation from our model predictions (predicted 1,914 $\pm$ 109; 876 fewer than observed). This increase is explainable by an increase in COVID-19 related research, as there were 864 quantitative biology pre-prints in 2020 with pandemic-related terms in their title or abstract (see Methods), and just 39 in the decade prior. Indeed, in {April 2020} nearly 60\% of the \arxiv\ pre-prints in quantitative biology were related to the pandemic (Figure~\ref{fig:2}). Pandemic-related pre-prints also appeared in other fields (computer science, physics, statistics, and quantitative finance, all peaking around {April 2020}), but these papers constituted fewer than 10\% of the total pre-prints in these fields. Approximately 10--15\% of current pre-prints in quantitative biology are related to the ongoing pandemic. This represents a strong shift in research focus, but this fraction is declining with time. % todo:

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date, \change{and that is before accounting for the tremendous work in scientific communication and contributions to public policy throughout the pandemic.} However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in {general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of \change{pre-prints} as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

We have taken a simple approach to address name confusion. We first define a unique author by family name and the initial of the first given name (`Family-name, I.'), such that we intentionally group together authors that may share the same initial of their first given name. 
While our approach to name confusion is grossly simple, it is unlikely that these choices have any substantial impact on our inferences. \change{Any author who changes their given or family name (or initials) will incorrectly appear as a `new' author when they next appear on a pre-print, but we expect this fraction to vary smoothly in time, allowing us to examine sharp changes as a consequence of the pandemic.} In general, \change{our approach} errs on the side of name conflation and over-productivity. 
Any common name is likely to appear in the literature early in the data set and will not impact the conclusions we draw about how publishing changed from 2020 January onwards. Indeed, we re-produced the analysis and figures in this paper by taking all initials given and it made no change to our conclusions. This is due, in part, to most authors not providing middle initials.

@@

We identified a pre-print as being related to the COVID-19 pandemic if the title or abstract contained any of the following four (case-insensitive) terms: `pandemic', `COVID', `SARS-CoV-2', `lockdown', \change{`spike glycoprotein`}, or `coronavirus'. Before 2020, these terms rarely appear in the title or abstract of any pre-print on \arxiv\ (Figure~\ref{fig:2}).

@@

\change{There were 234 pre-prints posted to the \texttt{q-bio} primary category in 2020 which include pandemic-related terms, and where no author had never appeared on the \texttt{q-bio} pre-print before. We investigated every pre-print to extract the affiliation of the lead author, and used this information to search the internet for the author's website. From this information we identified their research specialisations using their own words, or from their social media pages if no website could be found. In situations where an author's specialisation crossed multiple fields (e.g., mathematics and biology) we took a conservative approach and described the author as a biologist. However, we found that only 3\% (8/234) were led by established biologists who had not appeared in \texttt{q-bio} pre-prints before. About half (45\%; 107/234) were led by mathematicians or physicists, with another 26\% (61/234) led by engineers, computer scientists, or economists.}

@@

\bibitem{Davis:2007}
{Davis P. M., Fromerth M. J.,} {Does the arXiv lead to higher citations and reduced publisher downloads for mathematics articles?}, \emph{Scientometrics}, \textbf{71}, 2, doi:10.1007/s11192-007-1661-8 \newblock (2007).

\bibitem{GentilBeccot:2009}
{Gentil-Beccot A., Mele S., \& Brooks T.,} {Citing and Reading Behaviours in High-Energy Physics. How a Community Stopped Worrying about Journals and Learned to Love Repositories}, Preprint at https://arxiv.org/abs/0906.5418 \newblock (2009).
